%TODO: durchgehen und alles was hier exklusiv ist in die computation-kapitel aufnehmen

\chapter{Sequence Computations and Container Type}\indexmain{sequence computations}\label{seqcomp}

In this chapter we take a look at sequence computations, which are not concerned with directly controlling rules, but with computing values (sequence expressions) or causing side effects (sequence statements), that are then used to control rules.

\begin{example}
Sequence computations are typically employed when storages have to be maintained:
\verb#now:set<Node>=set<Node>{};>next:set<Node>=set<Node>{};>initializeNow(now) ;>#
\verb#( processNowFillNext(now, next) ;> { now.clear(); tmp:set<Node>=now; now=next;#
\verb#next=tmp; {!now.empty()} } )*#
--- that example sequence is used to implement a wavefront as rule control strategy.

A set \texttt{now} of current nodes is processed, filling a set of output nodes to be processed \texttt{next}, which are then used in the following iteration step as input nodes, until all of parts of the graph reachable from the initial nodes have been passed, yielding an empty set.
The sequence computation for switching in between the sets is given enclosed in braces in the sequence, 
the sequence expression to determine whether the wavefront came to a halt is given at the end of the computation,
enclosed a level deeper in braces.
As long as the set is still filled, the final expression, thus the computation, returns \texttt{true}, continuing with the loop -- when it gets empty, the expression yields \texttt{false}, henceforth terminating the loop and the wavefront algorithm.
\end{example}

Sequences are existing to control rules and not to carry out the bulk of the computational work,
so only a subset of the rule computations is available in the sequence computations.
The sequence computations are embedded directly in the controlling sequence on the other hand, in contrast to the computations from the rule language that can only be used by calling their named abstractions.
In particular, the subset that is needed in order to draw decisions (esp. comparison operators) and to assign variables (so for sequence variable initialization you need to employ a sequence computation) is available.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sequence Statements} \label{sec:seqcomp}

\begin{rail} 
  RewriteComputationUsage: (percent)? lbrace CompoundComputation rbrace; 
\end{rail}\ixnterm{RewriteFactor}

The non-computational sequence constructs introduced before in \ref{cha:xgrs} are used for executing rules, to determine which rule to execute next depending on success and failure of the previous rule applications, and where to apply it next by transmitting atomic variables of node or edge type in between the rules.
Sequence computations in contrast are used for manipulating container variables, evaluating computational expressions, or for causing side effects like output or element markings.
A computation returns always true, with exception of an expression used as a computation (explained below).
A prepended \texttt{\%} attaches a \indexed{break point} to the computation.

\begin{rail} 
  CompoundComputation: Computation ((';' Computation)*); 
\end{rail}

A compound computation consists of a computation followed by an optional list of computations separated by semicolons.
The computations are executed from left to right;
the value of the compound computation is the value of the last computation.
So you must give an expression at that point in order to return a value, 
whereas it is pointless to specify an expression before.
(But typically, when you are interested in a return value in a certain context, you only give a single computation, being an embraced sequence expression, which is then automatically the last computation.)

\begin{rail} 
  Computation:
     VariableDeclaration |
     Assignment |
     ProcedureMethodCall |
     ProcedureCall |
     lbrace SequenceExpression rbrace
  ;
	Assignment:	AssignmentTarget '=' (SequenceExpression | Assignment); 
\end{rail}\ixnterm{Computation}

A variable declaration declares a local variable in the same way as in the sequences.
An assignment assigns the value of a sequence expression to an assignment target.
It may be chained; such an assignment chain is executed from right to left, assigning the rightmost value to all the assignment targets given.
The expression used as computation -- denoted by and enclosed in braces -- will return a boolean value by comparing the return value of the expression to the default value of the corresponding type (e.g. 0 for an integer), returning false if equal, or true if not equal.
So just using a boolean variable as expression returns the value of the variable.
The form of expressions and assignment targets will be specified below.

\begin{note}
For returning an expression evaluation result value to a sequence you need two opening braces! 
One for entering the sequence computations, and the other for entering the sequence expressions.
\end{note}

\begin{example}
The example sequence
\verb#b:boolean ;> {b=true} ;> {{b}}#
shows a sequence with a variable declaration on sequence level, an assignment on sequence computation level (to initialize the variable), and a sequence expression defining the return value of the sequence to be the value of the variable.

The example sequence
\verb#if{ {{::i<10}}; {::i=::i+1}; false }*#
increments the graph global variable \texttt{i} until it becomes greater than \texttt{10}.
\end{example}

\begin{rail} 
	ProcedureMethodCall: ('(' Variables ')' '=')? \\ MethodTarget '.' ProcedureIdent '(' Arguments ')';
	MethodTarget: Variable | GraphElement '.' Attribute;
	ProcedureCall: ('(' Variables ')' '=')? ProcedureIdent '(' Arguments ')';
	Arguments: (SequenceExpression * ',');
	Varibales: (Variable + ',');
\end{rail}

A method call executes a method on a variable, passing further arguments.
The method may be one of the predefined container methods, or a user-defined method (cf. \ref{sec:objectoriented}).

A procedure call executes a user defined procedure from the rules file, cf. \ref{sub:compdef},
or a built-in procedure, passing further arguments.
The builtin procedures you may call here are e.g. \texttt{emit}, \texttt{emitdebug}, \texttt{record}, or \texttt{canonize(g:graph):string}, you find more on them in Chapter~\ref{cha:graph}.

Also procedures from packages may be called, the name \emph{ProcedureIdent} is then prefixed by the \emph{Ident} of the package, separated by a double colon \texttt{::}.
A user-defined procedure from a user-defined package may be called, or one of the built-in procedures contained in the built-in package \texttt{File}, e.g. \texttt{export}, or in the  built-in package \texttt{Transaction}, cf. Chapter~\ref{cha:graph}, or in the built-in package \texttt{Debug} (for more on packages see \ref{sub:packageaction}).

The \texttt{Debug} package procedures, as explained in detail in \ref{secdebuggersubrule}, are of interest in case sequences contained in rule execs are used, these are not directly visible to the debugger.
You call then \texttt{Debug::add} when entering and \texttt{Debug::rem} when leaving a subrule computation of interest (always pairwise!),
\texttt{Debug::emit} to recored some subrule computation milestones, \texttt{Debug::halt} to halt the debugger, and \texttt{Debug::highlight} to highlight some graph elements in the debugger (halting it).

Finally, an expression (without side effects) can be evaluated, this allows to return a (boolean) value from a computation.

\begin{rail}
  AssignmentTarget: 
    Variable (':' Type)? |
    'yield' Variable |
    GraphElement '.' Attribute |
    Variable '[' SequenceExpression ']' |
    GraphElement '.' Attribute '[' SequenceExpression ']' |
    GraphElement '.' 'visited' '[' SequenceExpression ']'
;
\end{rail}\ixnterm{AssignmentTarget}\ixkeyw{visited}\ixkeyw{yield}

Possible targets of assignments are the variables and def-variables to be yielded to, as in the simple assignments of the sequences. 
A \texttt{yield} assignment writes the rhs variable value to the lhs variable which must be declared as a  def-to-be-yielded-to variable (\texttt{def}-prefix) in the pattern containing the \texttt{exec} statement.
Yielding is only possible from compiled sequences, it always succeeds.
Further on, the attributes of graph elements may be written to, the values at given positions of array or deque or map variables may be written to, and the visited status of graph elements may be changed.

\begin{example}
The sequence computation \verb#{ x:int=42; y:N; (y)=proc(x); y.meth(x); y.a=x }# shows a variable declaration including an initialization (which falls out of scope at the closing brace), a variable declaration of node type without initialization, a call of a procedure with one input argument, assigning an output value, a call of a method of the node type, and the assignment of a graph element attribute.

The example \verb#{ x:array<int>=array<int>[]; x.add(42); x[0]=1; {x.size()>0} }# shows an array declaration and initialization, the adding of a value to the array, the indexed assignment to an array, and a terminal sequence expressions that makes the sequence computation succeed if the array is not empty (that's the case here) and fail otherwise. Sequence computations that don't end with a sequence expression always succeed.
\end{example}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sequence Expression} \label{sec:seqexpr}

\begin{rail}
  SequenceExpression:  
    ConditionalSequenceExpression |
    BooleanSequenceExpression |
    RelationalSequenceExpression |
    ArithmeticSequenceExpression |
    ContainerSequenceExpression |
    PrimarySequenceExpression;
\end{rail}\ixnterm{SequenceExpression}

Sequence expressions are a subset of the expressions introduced in \ref{sub:expr}, containing 
the boolean expressions,
the relational (comparison) expressions,
the arithmetic expressions, but not the bitwise expressions,
and the string expressions.
The container sequence expressions are a subset of the container expressions introduced in \ref{cha:container}, they will be explained in \ref{sec:storages}.

\begin{rail}
  ConditionalSequenceExpression: 
    BooleanSequenceExpression '?' SequenceExpression ':' SequenceExpression;
\end{rail}\ixnterm{ConditionalSequenceExpression}

The conditional operator has lowest priority, if the condition evaluates to true the first expression is evaluated and returned, otherwise the second.

\begin{rail}
  BooleanSequenceExpression: 
    SequenceExpression (ampersand | xorhat | '|' | doubleampersand | '||') SequenceExpression |
    '!' SequenceExpression;
\end{rail}\ixnterm{BooleanSequenceExpression}

The boolean operators have the same semantics and same priority as in \ref{sub:expr}.

\begin{rail}
  RelationalSequenceExpression: 
    SequenceExpression ('==' | '!=' | '<' | '<='| '>' | '>=' | titilde) SequenceExpression;
\end{rail}\ixnterm{RelationalSequenceExpression}

The equality operators work for every type and return whether the values to compare are equal or unequal.
The relational operators work as specified in \ref{sub:expr} for numerical types and \ref{cha:graph} for graph type.

\begin{rail}
  ArithmeticSequenceExpression:
    SequenceExpression ('+' | '-' | '*' | '/' | percent) SequenceExpression |
		'-' SequenceExpression;
\end{rail}\ixnterm{ArithmeticSequenceExpression}

The arithmetic operator plus is used to denote addition of numerical values or string concatenation,
the arithmetic operator minus is used to denote subtraction of numerical values.
Furthermore, you can multiply and divide numbers, or compute the remainder of a division.
Neither the arithmetic functions of package \texttt{Math}, nor the string methods are available in the sequence expressions.
Use the entities from the rule language for real computational work, you can reuse them easily, just call the functions from the sequence expressions, the procedures from the sequence statements, and the rules/tests from the sequences.

\begin{rail}
  PrimarySequenceExpression:
    BasicSequenceExpression |
    SpecialSequenceExpression;
\end{rail}\ixnterm{PrimarySequenceExpression}

The atoms of the sequence expressions are the basic and the special sequence expressions.

\begin{rail}
  BasicSequenceExpression:
    Literal |
    Variable | 
    GraphElement '.' Attribute |
    'def' '(' (Variable+',') ')' |
    railat '(' NameString ')'
  ;
\end{rail}\ixnterm{BasicSequenceExpression}\ixkeyw{def}

The basic sequence expressions are the foundational value sources.
The literal and variable basic expressions are the same as in the \emph{SimpleOrInteractiveExpression}, cf. \ref{sec:simplevarhandling}, or as in the \emph{PrimaryExpression}s of the rule language, cf. \ref{sec:primexpr},
this means esp. that a Variable may denote a graph global variable if prefixed with a double colon, here as well as in the \emph{AssignmentTarget}.
The attribute access clause returns the attribute value of the given graph element.
A \texttt{def} term is successful iff all the variables are defined (not null).
The at operator allows to access a graph element by its \indexed{persistent name}.

\begin{rail}
  SpecialSequenceExpression:
    Variable '[' SequenceExpression ']' |
    GraphElement '.' Attribute '[' SequenceExpression ']' |
    GraphElement '.' 'visited' '[' SequenceExpression ']' |
    FunctionMethodCall |
    FunctionCall |
    RuleQuery |
    MultiRuleQuery
  ;
\end{rail}\ixnterm{SpecialSequenceExpression}\ixkeyw{visited}

The special sequence expressions consist of indexed accesses, function method calls, function calls, and rule queries.

They are esp. used for storage handling, and for graph, subgraph and visited flag handling.
The storage oriented ones are used to access a storage or to call a method on a storage. 
They were introduced in chapter \ref{cha:container}, and are summarized below in \ref{sec:storages}.
The graph and subgraph handling expressions allow to query the graph for its elements, the visited flags expressions allow to check whether a value is marked.
They were introduced in chapter \ref{cha:graph}, and are summarized below in \ref{sec:queryupdate}.

\begin{rail} 
	FunctionMethodCall: MethodTarget MethodSelector;
	MethodSelector: '.' FunctionIdent '(' Arguments ')' (MethodSelector)?;
	FunctionCall: FunctionIdent '(' Arguments ')';
\end{rail}

A method call executes a method on a variable, passing further arguments.
The method may be one of the predefined container methods, or a user-defined method (cf. \ref{sec:objectoriented}).
Method calls work as in the rule language, i.e. you can build chains of function method calls, like with the \emph{MethodSelector} from the rule language, cf. Chapter~\ref{cha:typeexpr} (in contrast to procedure method calls).

A function call executes a user defined function from the rules file, cf. \ref{sub:compdef},
or a built-in function, passing further arguments.

An available built-in function you may call here is e.g. \texttt{random} that behaves like the random function from the expressions, see \ref{sec:primexpr}.

Further available are \texttt{nameof} to fetch the name of a node or edge or graph and \texttt{uniqueof} to fetch the unique id of a node or edge or graph.
The function \texttt{nodeByName} allows to retrieve a node by its name, \texttt{edgeByName} does the same for an edge.
The function \texttt{nodeByUnique} allows to retrieve a node by its unique id, \texttt{edgeByUnique} does the same for an edge.
You find more on them in Chapter~\ref{cha:performance}.

The \texttt{typeof} function returns the type \emph{as string}, for an arbitrary \GrG-object fed as input.
There is no type type supporting type comparisons as in the rule language existing, you are limited to string comparisons.

Also functions from packages may be called, the name \emph{FunctionIdent} is then prefixed by the \emph{Ident} of the package, separated by a double colon \texttt{::}.
A user-defined function from a user-defined package may be called, or one of the built-in functions contained in the built-in package \texttt{File}, e.g. \texttt{import}, or \texttt{now} from the built-in package \texttt{Time}, cf. Chapter~\ref{cha:graph} (for more on packages see \ref{sub:packageaction}).

Regarding the rule queries, take a look at Chapter~\ref{cha:graphquery}.

\begin{example}
The sequence expression \verb#{{ def(y) && y.meth(x)>=y.a || @("$1").a+1 != func(x) }}# checks whether the node typed variable \texttt{y} is defined, i.e. not null, and if so compares the return value of a method of that type with an attribute value. If the comparison succeeds, it defines the return value of the expression; in case the definedness check failed or the comparison yields false, is a graph element fetched from the graph by its persistent name \verb#$1#, and its attribute \texttt{a} plus 1 compared against the result of a function call. Then the result of this latter comparison defines the outcome of the expression.
\end{example}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Storages -- Container Type Handling in the Sequences}\label{sec:storages}\indexmain{storage}
Storages are variables of container (set/map/array/deque) type (cf. \ref{sec:builtingenerictypes}) storing nodes or edges.
They are primarily used in the sequences, from where they are handed in to the rules via \texttt{ref} parameters (but additionally container attributes in graph elements may be used as storages,
esp. for carrying out data flow analyses, cf. \ref{subsub:flow}).
They allow to decouple processing phases: the first run collects all graph elements relevant for the second run which consists of a sequence executed for each graph element in the container.
The splitting of transformations into passes mediated by container valued global variables allows for subgraph copying without model pollution, cf. \ref{subsub:copystructure}; please have a look at \ref{sub:mergesplit}, \ref{subsub:copystructure} and \ref{subsub:flow} regarding a discussion on when to use which transformation combinators and for storage examples.
They were already defined and described in \ref{cha:container}.
Here we only give some refinements and explanations of the semantics.

\subsubsection*{Common methods and operators}

The update methods \texttt{add}, \texttt{rem}, and \texttt{clear} are available for all storages and allow to add elements to the container, remove elements from the container, or clear the container.
Their return value is the changed container, thus they allow to chain method calls on the container.

The query methods \texttt{size}, \texttt{empty}, \texttt{peek} in contrast return the size of the container, whether the container is empty, or a certain element from the container and thus can't be chained.

The indexed query \texttt{v=m[k]} is available on map and array and deque types and returns the element at the specified index,
the indexed assignment \texttt{a[i]=v} overwrites the element at the specified index.

\begin{rail}
  ContainerSequenceExpression: 
    SequenceExpression ('|' | ampersand | backslash | '+') SequenceExpression;
\end{rail}\ixnterm{ContainerSequenceExpression}

All infix container operators are available in the sequence expressions, they have the same semantics and same priority as in Chapter~\ref{cha:container}.

\begin{rail}
  RelationalSequenceExpression: 
    SequenceExpression ('==' | '!=' | '<' | '<='| '>' | '>=' | 'in') SequenceExpression;
\end{rail}\ixnterm{RelationalSequenceExpression}

The relational (including equality) operators work as specified in Chapter~\ref{cha:container} for container types.
Further available is the sequence expression operator \texttt{in} for membership query -- it is not supported on string type, though.

All container constructors and copy constructors are supported, see Chapter~\ref{cha:container} for more on them.
The change assignments known from rule language in contrast are \emph{not} available in the sequence computations.

Further support is discussed in the following separated by type.

\subsubsection*{Array Operations}
All array methods are supported (the array of match is the central data structure in the sequence expressions to support pattern-based graph queries), see Section~\ref{sec:arrayexpr} and Section~\ref{sec:accumulation} for a list of them.
From the array methods specifying and accessing a member of the array value type, the \texttt{extract<member>()} method is supported as well as the ordering methods \texttt{order\-Ascending\-By<member>()} and \texttt{order\-Descending\-By<member>()}, plus \texttt{group\-By<member>()} and \texttt{keep\-One\-For\-Each<member>()} (note that further means of ordering are available as filter calls).

All infix operators are supported, which are array concatenation, as well as the relational operators.

\subsubsection*{Set Operations}
All set methods are supported, this is in addition to the ones already mentioned in the section on the common parts of all container types: \texttt{asArray()} (cf. \ref{sec:setexpr}).

All infix operators are supported, which are set union, intersection, and difference, as well as the relational operators.

\subsubsection*{Map Operations}
All map methods are supported, these are in addition to the ones already mentioned in the section on the common parts of all container types: \texttt{domain()}, \texttt{range()}, and \texttt{asArray()} (cf. \ref{sec:mapexpr}).

All infix operators are supported, which are map union, intersection, and difference (including sets as right value), as well as the relational operators.

\subsubsection*{Deque Operations}
All deque methods are supported, these are in addition to the ones already mentioned in the section on the common parts of all container types: \texttt{subdeque()}, \texttt{asArray()}, \texttt{asSet()}, \texttt{indexOf(.), \texttt{indexOf(.,.)}}, \texttt{lastIndexOf(.)} (cf. \ref{sec:dequeexpr}).

All infix operators are supported, which are deque concatenation, as well as the relational operators.

\subsubsection*{Container For Loop}

You may iterate in the sequences with a for loop over the elements contained in a storage.

\begin{rail}
  RewriteFactor:
    'for' lbrace (Var 'in' SetVar | Var '->' Var 'in' MapVar | Var 'in' ArrayVar | Var '->' Var 'in' ArrayVar | Var 'in' DequeVar | Var '->' Var 'in' DequeVar) ';' RewriteSequence rbrace
    ;
\end{rail}\ixkeyw{in}\ixkeyw{for}\ixnterm{RewriteFactor}\label{forstorage}

The \texttt{for} command iterates over all elements in the set or array or deque, or all key-value pairs in the map or array or deque, and executes for each element / key-value pair the nested graph rewrite sequence; it completes successfully iff all sequences were executed successfully (an empty container causes immediate successful completion); the key in the key-value pair iteration of an array or deque is the integer typed index. (See \ref{forgraphelem} for another version of the \texttt{for} command.)

\begin{example}
The following XGRS is a typical storage usage.
First an empty set \texttt{x} is created, which gets populated by a rule \texttt{t} executed iteratedly, returning a node which is written to the set.
Then another rule is executed iterated for every member of the set doing the main work, and finally the set gets cleared to prevent memory leaks or later mistakes.
If the graph should stay untouched during set filling you may need \texttt{visited} flags to prevent endless looping.
\verb#x=set<Node>{} ;> ( (v)=t() && {x.add(v)} )+ && for{v in x; r(v)} <; {x.clear()}#
You could hand in the storage to the rule, and \texttt{add} there to the set, this would allow to shorten the sequence to:\\
\verb#x=set<Node>{} ;> ( t(x) )+ && for{v in x; r(v)} <; {x.clear()}#\\
The for loop could be replaced by employing the storage access in the rule construct, cf. \ref{sub:storageaccess}; this would be especially beneficial if the rule \texttt{r} inside the for loop would have to change the storage \texttt{x}, which would corrupt the iteration/enumeration variable.
\end{example}

\begin{warning}
The container over which the \texttt{for} loop iterates must stay untouched during iteration.

Use the \texttt{copy(.:container):container} function to clone the container before the iteration if you need to iterate the container content \emph{and} change its values.
\end{warning}

%\begin{example}
%\begin{grgen}
%::x=set<NodeTypeA>{}
%y:map<Node,Edge> = map<Node,Edge>{u->v}
%\end{grgen}
%The first line declares or references a global variable \texttt{x} (without static type) and assigns the newly created, empty set of type \texttt{set<NodeTypeA>} to it as value.
%The second line declares a variable \texttt{y} of type \texttt{map<Node,Edge>} and assigns the newly created map of the same type containing a key value pair build from u mapping to v, where u is assumed to be a variable of node type and v to be a variable of edge type.
%\end{example}

%\pagebreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Graph and Subgraph Based Queries and Updates}\label{sec:queryupdate}\label{sec:visited}

The graph and subgraph oriented parts of the sequence expressions are built from four groups:
\begin{enumerate}
	\item the procedures for basic graph manipulation,
	\item the functions for querying the graph structure,
	\item the functions and procedures of the subgraph operations, and
	\item the visited flag query, assignment, and procedures.
\end{enumerate}

The first group is built from basic graph manipulation operators, as defined in \ref{procstab} and described in \ref{sub:procedures}.
Elements may be added, removed, or retyped, and nodes may be merged or edges redirected.
Not available are numerical functions, they are only offered by the computations of the rules. 

\begin{example}
The sequence computation \verb#{ (::x)=add(N); (::x)=retype(::x,M); rem(::x) }# adds a newly created node of type \texttt{N} to the graph (storing it in the global variable \verb#::x#), retypes it to \texttt{M}, and finally removes it again from the graph.
\end{example}

The second group is built from the operators querying primarily the connectedness of graph elements,
as defined in \ref{sub:functions}.
You may ask for one for the nodes or edges of a type.
You may query for the other for the source or target or opposite node of an edge.
Furthermore, you may query for adjacent nodes and incident edges,
maybe even transitively for the reachability.
Furthermore you may ask with a predicate whether nodes or edges are adjacent or incident to other nodes or edges, maybe even transitively for reachability.

These global functions returning elementary data define the simple way of graph querying -- the more powerful way is defined by pattern based queries (employing rules/tests) -- they return an array of pattern matches, see Chapter~\ref{cha:graphquery}.

\begin{example}
\verb#for{x:N in nodes(N); for{::y in outgoing(x); { ::z=target(::y); ::z.a = 42 } } }# 
is a sequences that sets the attribute \texttt{a} to 42 for all nodes that are adjacent as targets to a source node \texttt{x} of type \texttt{N}.
You will receive a runtime exception if the type of \verb#::z# does not possess an attribute \texttt{a}.

A more realistic example is to check whether two nodes returned by some rule applications are reachable from each other, carrying out a change only in this case:\\
\verb#(::x)=r() ;> (::y)=s() ;> if{ {{isReachable(::x,::y)}} ; doSomething(::x, ::y) }#
\end{example}

The third group is defined by functions and procedures that operate on (sub-)graphs, as defined in \ref{sub:functions} and \ref{sub:procedures}.
They are especially useful in state space enumeration, cf. \ref{sec:statespaceenum}.
To this end, parallelized graph isomorphy checking with the \texttt{equalsAny} function is especially of interest.
You may import, clone, or compute induced subgraphs.
You may export a subgraph or insert a subgraph into the hostgraph.

\begin{example}
\verb#( doSomething() ;> { File::export("graph"+i+".grs") } )*# is a sequence scheme for exporting a graph after each iteration step in a loop, gaining a series of snapshots on the hard drive.
In a later step, you may then conditionally add exported graphs to the host graph: 
\verb#if{cond; { ::g=File::import("graph"+n.a+".grs"); insert(::g) } }#
\end{example}

\begin{example}
When you model a state space with \texttt{Graph} representative nodes (\emph{not} \texttt{graph} standing for a real (sub)graph), which are pointing with \texttt{contains} edges to the nodes contained in their state (i.e. subgraph), and store additionally a replica of the subgraph in a \texttt{sub} attribute of the \texttt{Graph} node, so it is readily available for comparisons,
then the step of a state space enumeration with isomorphic state pruning is controlled with code like this:\\
\verb#<< modifyCurrent(gr) ;; {adj=adjacent(gr, contains); sub=inducedSubgraph(adj)}#\\
Inside the backtracking double angles, a new state is computed as first step by modifying the currently focused state received as input \texttt{gr:Graph} from the previous step. The modified subgraph is extracted for comparison by computing the \texttt{inducedSubgraph} from the nodes \texttt{adjacent} via \texttt{contains}-edges to the \texttt{gr}-node.\\
\verb#;> for{others:Graph in nodes(Graph); {{sub!=others.sub}} } && #\\
The extracted subgraph is compared with all already enumerated subgraphs that can be accessed by their \texttt{Graph} representative node. Only if none is isomorphic to it, do we continue with making the state persistent.\\
\verb#/ {(ngr)=insertInduced(adj, gr)} && link(gr,ngr) && {ngr.sub=sub} /#\\
During a backtracking pause, the modified subgraph is cloned and inserted flatly into the host graph again with \texttt{insertInduced}. A link is added from the old representative to this new representative, to reflect ancestry. Then the subgraph attribute of the new representative \texttt{ngr} is filled with the previously computed subgraph \texttt{sub}. Remark: the first \texttt{inducedSubgraph} above does not contain the representative node and thus is missing all containment edges, too. This \texttt{insertInduced} includes the representative node and thus the containment edges. Syntactical remark: \texttt{inducedSubgraph} is used in an assignment with a function call as RHS, whereas \texttt{insertInduced} is employed from a procedure call which requires parenthesis around the output arguments.\\
\verb#&& stateStep(ngr, level+1) >>#\\
Finally, we continue state space construction with the next step, modifying the just inserted subgraph.
After this step returns (with \texttt{false} as result), do the backtracking double angles roll back the modification -- keeping the changes written during the pause untouched -- and execute \texttt{modifyCurrent} on the next match available in \texttt{gr}.
\end{example}

The fourth group are the visited flags related operations,
as described in chapter \ref{sub:visitedaccess}.
Available is an expression for reading a visited flag, an assignment for writing a visited flag, and procedures for managing the visited flags as defined in \ref{procstab}.

\begin{example}
Because of the need to allocate and deallocate them, the visited flags are typically used with code like this:
\verb#flag:int ;> {(flag)=valloc()} ;> r(flag) ;> {vfree(flag)}#\\
In addition, they may be read in the sequence expressions, and written in the sequence computations:
\verb#if{ {{!n.visited[flag]}} ; { n.visited[flag] = true } }#
\end{example}

In the sequences only the sequence expressions are available to compute the parameters for the functions and procedures, compared to the full-fledged expressions of the computations language.

%\pagebreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Quick Reference Table}

Table~\ref{comptab} lists most of the operations of the graph rewrite computations at a glance.

 %\makeatletter
\begin{table}[htbp]
\begin{minipage}{\linewidth} \renewcommand{\footnoterule}{} 
\begin{tabularx}{\linewidth}{|lX|}
\hline
\texttt{c;d} & Computes c then d; the value of the computation is d\\
\hline
\texttt{\{e\}} & An unspecified sequence expression executed for its result, defining the success of the computation.\\
\hline
\texttt{t=e} & Simple assignment of an expression value to an assignment target\\
%\texttt{t=e=f} & Chained assignment \\
\hline
\texttt{e ? f : g} & Returns f if e evaluates to true, otherwise g \\
\texttt{e op f} & For \texttt{op} being one of the boolean operators \texttt{||,|,\&,\&\&,\^\ } \\
\texttt{e op f} & For \texttt{op} being one of comparison operators \texttt{==,!=,<,<=,>,>=,in} \\
\texttt{e op f} & For \texttt{op} being one of arithmetic operators \texttt{+,-,*,/,\%} \\
\texttt{e op f} & For \texttt{op} being one of the set or map operators \texttt{|,\&,\textbackslash} \\
\texttt{e + f} & For array or deque or string concatenation. \\
\texttt{op e} & For \texttt{op} being boolean \texttt{!} or arithmetic \texttt{-}\\
\hline
\texttt{[?r\textbackslash f]} & Rule query, returns every match produced by \texttt{r} that was left by the filter \texttt{f}, in the order left behind by the filter, as \texttt{array<match<r>>}. \\
\texttt{[?[q,r]\textbackslash mc.f\textbackslash<mc>]} & Multi rule query, returns every match produced by \texttt{q} and \texttt{r} that was left by the filter \texttt{f} of match class \texttt{mc}, in the order left behind by the match class filter, as \texttt{array<match<class mc>>}. \\
\hline
\texttt{v} & Variable. Assignment target or expression.\\
\texttt{v.name} & Attribute of graph element. Assignment target or expression.\\
\texttt{@(name)} & Return graph element of given name.\\
\texttt{def(\emph{Parameters})} & Check if all the variables are defined.\\
\texttt{random(upperBound)} & Returns random number from [0;upper bound[, if upper bound is missing from [0.0;1.0[.\\
\texttt{typeof(v)} & Returns the name of the type of the entity handed in.\\
\hline
\texttt{u=set<Node>\{\}} & Example for container constructor, creates storage set and assigns to \texttt{u}.\\
\texttt{u[e]} & Target value of \texttt{e} in \texttt{u}. Fails if \texttt{!(e in u)}. Assignment target or expression.\\
\hline
\texttt{f(...)} & Calls one of the functions for graph querying defined in \ref{funcstab} and explained in \ref{cha:graph}. Or calls a user defined function. The numerical functions are \emph{not} available.\\
\hline
\texttt{(...)=p(...)} & Calls one of the procedures for graph manipulation defined in \ref{procstab} and explained in \ref{cha:graph}. Or calls a user defined procedure.\\
\hline
\texttt{v.fm(...)} & Calls one of the function methods \texttt{size,empty,peek} for container querying defined in \ref{funcmethstab} and explained in \ref{cha:container}. Or calls a user defined function. The string function methods and other container function methods as well as infix operators are \emph{not} available.\\
\hline
\texttt{(...)=v.pm(...)} & Calls one of the procedure methods \texttt{add,rem,clear} for container manipulation defined in \ref{procmethstab} and explained in \ref{cha:container}. Or calls a user defined procedure. The change assignments are \emph{not} available.\\
\hline
\end{tabularx}
\end{minipage}\\
\\ 
{\small Let \texttt{c} and \texttt{d} be computations, \texttt{t} be an assignment target, \texttt{e}, \texttt{f}, \texttt{g} be expressions, \texttt{u}, \texttt{v}, \texttt{w} be variable identifiers, \texttt{q} and \texttt{r} be rules, \texttt{mc} be a match class, \texttt{f} be a filter }
\caption{Sequence computations at a glance}
\label{comptab}
\end{table}
%\makeatother
 
% todo: beispiele im text bringen


